Mastery Learning in Practice:\\ A Descriptive Analysis of Data from the Cognitive Tutor Algebra I Effectiveness Trial

### 2017-10-19 paper.Rnw
sss

@@

\begin{figure}
<<transition1,fig.height=3,fig.width=6.5,dependson='secOrder'>>=

trans <- with(filter(secOrderCP,year==1),table(unitName,nextUnitName))

## top senders and receivers
cols <- which(colSums(trans)>30)
rows <- which(rowSums(trans)>30)

@@

## top 2 receivers for each sender, vice versa
for(r in rows)
    while(sum(trans[r,cols2])/sum(trans[r,])<0.75){
    nextBest <- which.max(trans[r,seq(ncol(trans))[-cols2]])
    cols2 <- c(cols2,which(colnames(trans)==names(nextBest)))
}

for(cc in cols){
 #   print('new col')
  #  print(cc)
    while(sum(trans[rows2,cc])/sum(trans[,cc])<0.75){
        nextBest <- which.max(trans[seq(nrow(trans))[-rows2],cc])
        rows2 <- c(rows2,which(rownames(trans)==names(nextBest)))
    }

@@

trans <- trans[rows,cols]
## trans <- trans[rowSums(trans)>50,colSums(trans)>10]
## trans <- trans[rowSums(trans)>50,colSums(trans)>10]
## trans <- trans[rowSums(trans)>50,colSums(trans)>10]

propTotCP <- sum(trans)/sum(secOrderCP$year==1)
#transitionPlot(trans)

nsec <- nrow(trans)
uuu <- levels(secOrderCP$unitName)
uuu <- uuu[uuu%in%c(colnames(trans),rownames(trans))]
nsec <- length(uuu)

@@

for(i in 1:nsec) if(uuu[i]%in%rownames(trans)) text(1,nsec-i+1,uuu[i],adj=c(1,NA))
for(i in 1:nsec) if(uuu[i]%in%colnames(trans)) text(2,nsec-i+1,uuu[i],adj=c(0,NA))

for(rr in rownames(trans))
 for(cc in colnames(trans)){
#     if(trans[rr,cc]>20){
      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans[rr,cc]/sum(trans[rr,]),
       col=rgb(0,0,0,alpha=ifelse(trans[rr,cc]>50,1,trans[rr,cc]/50)))
}

@@

trans <- with(filter(secOrderCP,year==2),table(unitName,nextUnitName))

## top senders and receivers
cols <- which(colSums(trans)>30)
rows <- which(rowSums(trans)>30)

@@

## top 2 receivers for each sender, vice versa
for(r in rows)
    while(sum(trans[r,cols2])/sum(trans[r,])<0.75){
    nextBest <- which.max(trans[r,seq(ncol(trans))[-cols2]])
    cols2 <- c(cols2,which(colnames(trans)==names(nextBest)))
}

for(cc in cols){
 #   print('new col')
  #  print(cc)
    while(sum(trans[rows2,cc])/sum(trans[,cc])<0.75){
        nextBest <- which.max(trans[seq(nrow(trans))[-rows2],cc])
        rows2 <- c(rows2,which(rownames(trans)==names(nextBest)))
    }

@@

trans <- trans[rows,cols]
## trans <- trans[rowSums(trans)>50,colSums(trans)>10]
## trans <- trans[rowSums(trans)>50,colSums(trans)>10]
## trans <- trans[rowSums(trans)>50,colSums(trans)>10]

propTotCP <- sum(trans)/sum(secOrderCP$year==1)
#transitionPlot(trans)

nsec <- nrow(trans)
uuu <- levels(secOrderCP$unitName)
uuu <- uuu[uuu%in%c(colnames(trans),rownames(trans))]
nsec <- length(uuu)

@@

for(i in 1:nsec) if(uuu[i]%in%rownames(trans)) text(1,nsec-i+1,uuu[i],adj=c(1,NA))
for(i in 1:nsec) if(uuu[i]%in%colnames(trans)) text(2,nsec-i+1,uuu[i],adj=c(0,NA))

for(rr in rownames(trans))
 for(cc in colnames(trans)){
#     if(trans[rr,cc]>20){
      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans[rr,cc]/sum(trans[rr,]),
       col=rgb(0,0,0,alpha=ifelse(trans[rr,cc]>50,1,trans[rr,cc]/50)))
}

>>>>

Teachers who reassign students may simply move them to the next
section within the same unit.
Say that a teacher believes that a particular student had already mastered
the skills in one of the CTAI sections, or is wheel-spinning---working
problems without learning.
Still, the teacher wants the student to learn as much as possible from
the student's current unit.
Then moving the student to the next section within the same unit would
make sense.
Alternatively, teachers who believe that some of their students are
not progressing quickly enough may reassign them out of their current
units entirely, and into the next unit in the sequence---whatever that
may be.
Finally, a teacher who wanted his or her students to focus on a
particular topic might reassign them all to the appropriate section
when the time is appropriate.

Which of these patterns is most prevelant?
More generally, when teachers reassign their students, where in the
curriculum do they send them?

To address these questions, we focus on the units in the standard
Algebra I sequence.
Figures \ref{fig:trans1} and \ref{fig:trans2} give transition plots
for reassignment in the two study years.
On the left of each figure, marked ``From,'' are the top ``sending''
units: the units that students were reassigned \emph{from} at least 30
times.
They are listed in the order they appear in the standard Algebra I curriculum.
On the right, under ``To,'' are the top ``receiving'' units: the units
that students were reassigned \emph{to} at least 30 times.
For completeness, each of the top two receivers for each sending unit,
and each of the top two senders for each receiving unit were also included.
Finally, all of the sending units were also listed on the left, in the
receiving column.

The arrows in the plot represent reassignments.
An arrow from a sending unit to a receiving unit indicates
reassignment from the former to the latter.
The thickness of the arrows represents the proportion of reassignments
originating in the sending unit whose destination was the receiving unit.
The darkness of the arrows represents the number of such reassignments.
For instance, in Figure \ref{fig:trans1}, the thick line from
``4−Quadrant Linear Graphs'' on the left to ``4−Quadrant Linear
Graphs'' on the right is fairly thick, since

@@

\begin{figure}
  \begin{subfigure}{.7\textwidth}
<<transition1,fig.height=3,fig.width=4.5,dependson='secOrder'>>=

trans1 <- with(filter(secOrderCP,year==1),table(unitName,nextUnitName))

## top senders and receivers
cols <- which(colSums(trans1)>30)
rows <- which(rowSums(trans1)>30)

@@

## top 2 receivers for each sender, vice versa
for(r in rows)
    while(sum(trans1[r,cols2])/sum(trans1[r,])<0.75){
    nextBest <- which.max(trans1[r,seq(ncol(trans1))[-cols2]])
    cols2 <- c(cols2,which(colnames(trans1)==names(nextBest)))
}

for(cc in cols){
    while(sum(trans1[rows2,cc])/sum(trans1[,cc])<0.75){
        nextBest <- which.max(trans1[seq(nrow(trans1))[-rows2],cc])
        rows2 <- c(rows2,which(rownames(trans1)==names(nextBest)))
    }

@@

trans1 <- trans1[rows,cols]

propTotCP <- sum(trans1)/sum(secOrderCP$year==1)
#trans1itionPlot(trans1)

nsec <- nrow(trans1)
uuu <- levels(secOrderCP$unitName)
uuu <- uuu[uuu%in%c(colnames(trans1),rownames(trans1))]
nsec <- length(uuu)

@@

for(i in 1:nsec) if(uuu[i]%in%rownames(trans1)) text(1,nsec-i+1,uuu[i],adj=c(1,NA))
for(i in 1:nsec) if(uuu[i]%in%colnames(trans1)) text(2,nsec-i+1,uuu[i],adj=c(0,NA))

for(rr in rownames(trans1))
 for(cc in colnames(trans1)){
#     if(trans1[rr,cc]>20){
      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans1[rr,cc]/sum(trans1[rr,]),
       col=rgb(0,0,0,alpha=ifelse(trans1[rr,cc]>50,1,trans1[rr,cc]/50)))
}

@
\end{subfigure}
\begin{subfigure}{.25\textwidth}
<<trans1legend,fig.width=1.6,fig.height=3>>=
par(mar=c(.1,.1,.1,.1))
plot(1,1,xlim=c(0,2),ylim=c(0,11),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
text(1,10,expression(underline(bold(paste(Legend)))))
text(1,8,'Thickness: Percent')
text(.5,7,'25%')
text(.5,6,'50%')
text(.5,5,'75%')
arrows(c(1.5,1.5,1.5),c(7,6,5),c(2,2,2),lwd=5*c(.25,.5,.75))
text(1,4,'Darkness: Number')
text(.5,3,'10')
text(.5,2,'25')
text(.5,1,'50+')
arrows(c(1.5,1.5,1.5),c(3,2,1),c(2,2,2),col=rgb(0,0,0,alpha=c(1/5,.5,1)))

@
\caption{Reassignment transition plot for study year 1. The units on
  the left of the plot are the Algebra I units that ended in reassignment at
  least 30 times in year 1. The units on the right are those that were
  the destination of at least 30 reassignments. Also included on the
  left are the top two sending units for each of the units on the
  right, and also included on the right are all the units on the left,
  along with their top two receiving units. The units are listed
  following their order in the standard Algebra I curriculum. There is an arrow from a
  unit on the left to a unit on the right if a student was assigned
  from the unit on the left to the one on the right. The thickness of
  the arrows is proportional to the percentage of all reassignments from
  the sending unit that ended in that receiving unit. The darkness of
  the arrows is proportional to the number of reassignments from the
  sending unit to the receiving unit.}
\label{fig:trans1}
\end{figure}

@@

trans2 <- with(filter(secOrderCP,year==2),table(unitName,nextUnitName))

## top senders and receivers
cols <- which(colSums(trans2)>30)
rows <- which(rowSums(trans2)>30)

@@

## top 2 receivers for each sender, vice versa
for(r in rows)
    while(sum(trans2[r,cols2])/sum(trans2[r,])<0.75){
    nextBest <- which.max(trans2[r,seq(ncol(trans2))[-cols2]])
    cols2 <- c(cols2,which(colnames(trans2)==names(nextBest)))
}

for(cc in cols){
 #   print('new col')
  #  print(cc)
    while(sum(trans2[rows2,cc])/sum(trans2[,cc])<0.75){
        nextBest <- which.max(trans2[seq(nrow(trans2))[-rows2],cc])
        rows2 <- c(rows2,which(rownames(trans2)==names(nextBest)))
    }

@@

trans2 <- trans2[rows,cols]

propTotCP <- sum(trans2)/sum(secOrderCP$year==1)

nsec <- nrow(trans2)
uuu <- levels(secOrderCP$unitName)
uuu <- uuu[uuu%in%c(colnames(trans2),rownames(trans2))]
nsec <- length(uuu)

@@

for(i in 1:nsec) if(uuu[i]%in%rownames(trans2)) text(1,nsec-i+1,uuu[i],adj=c(1,NA))
for(i in 1:nsec) if(uuu[i]%in%colnames(trans2)) text(2,nsec-i+1,uuu[i],adj=c(0,NA))

for(rr in rownames(trans2))
 for(cc in colnames(trans2)){
#     if(trans2[rr,cc]>20){
      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans2[rr,cc]/sum(trans2[rr,]),
       col=rgb(0,0,0,alpha=ifelse(trans2[rr,cc]>50,1,trans2[rr,cc]/50)))
}
@
\caption{Reassignment transition plot for study year 2.}
\label{fig:trans2}
\end{figure}
